---
title: "STAT40830 - Homework 1"
author: "Ashish Mohamed Usman 24201306"
format: pdf
execute:
  echo: false
---

# *Introduction*

This analysis presents a relationship between the **number of drivers killed** and time, before and after seatbelt legislation was introduced in the UK. The legislation came into effect in **January 1983**, and its impact can be observed through changes in the number of monthly road casualties.

The R built-in dataset used in this report is **Seatbelts** from the `datasets` package. It contains monthly UK road casualty statistics from **January 1969 to December 1984**, across various categories.

## ***Understanding Dataset :***

```{r}
# Loading the Seatbelts dataset
data(Seatbelts)

# Displaying the structure of the dataset
# This shows the number of observations, variables, and data type
str(Seatbelts)
```

The `Seatbelts` dataset has **192 monthly observations** and includes the following variables:

```{r}
tail(Seatbelts)
```

**DriversKilled** : Car drivers killed\
**drivers** : Car drivers killed or seriously injured\
**front** : Front-seat passengers killed or seriously injured\
**rear** : Rear-seat passengers killed or seriously injured\
**kms** : Distance driven (km)\
**PetrolPrice** : Petrol price\
**VanKilled** : Van driver casualties\
**law** : 1 if seatbelt law is in effect, 0 otherwise

## *Data Summary :*

Below is a descriptive summary of the main variable - 'DriversKilled' used in the graph:

```{r}
#| echo: false

# Showing summary statistics for the number of drivers killed per month
summary(Seatbelts[, "DriversKilled"])

```

The number of drivers killed per month ranges from a **minimum of 60** to a **maximum of 198**, with a **median of 118.5**, indicating that half of the observations fall below this value.

The **mean is 122.8**, which is slightly higher than the median, suggesting a slight **right-skew** in the distribution.

The **1st quartile** is 104.8 and the **3rd quartile** is 138.0, showing that 50% of the values lie within this range.

## *Plot :*

The following plot displays the number of **drivers killed per month**, with a red dashed line indicating the start of the seatbelt law.

```{r}
# Loading ggplot2 package for plotting
library(ggplot2)

# Converting the Seatbelts time series to a data frame with date and variables
df <- data.frame(
  # Monthly dates
  date = seq(as.Date("1969-01-01"), as.Date("1984-12-01"), by = "month"), 
  # Number of drivers killed each month
  DriversKilled = as.numeric(Seatbelts[, "DriversKilled"]),   
  # 0 or 1 indicating if seatbelt law was in effect
  law = as.numeric(Seatbelts[, "law"])                                     
)

# Create the plot
ggplot(df, aes(x = date, y = DriversKilled)) +
   # Line plot in dark blue
  geom_line(color = "yellow", linewidth = 1) +      
  # Red dashed line for seatbelt law start
  geom_vline(xintercept = as.Date("1983-01-01"), linetype = "dashed", 
             color = "red") +                                              
  labs(
    title = "Monthly UK Driver Deaths (1969–1984)",   # Title of the plot                     
    x = "Date", # X-axis label                                                           
    y = "Drivers Killed"  # Y-axis label                                                 
  ) +
  theme_minimal()                                                         

```

## *Description of the Plot :*

The time series plot illustrates the **monthly number of drivers killed in the UK from 1969 to 1984**. A clear **seasonal pattern** is visible, with regular peaks and troughs occurring each year, typically higher in certain months, likely corresponding to increased travel periods.

A **red dashed vertical line** marks **January 1983**, when **seatbelt legislation** was introduced. Following this date, there is a **notable decrease in fatalities**, suggesting that the implementation of mandatory seatbelt use had a **positive impact on road safety**, reducing the number of driver deaths in subsequent months.
